---
title: "EDA"
author: "Amruta Gumma"
date: "16 September 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```


## R Markdown



```{r , include = FALSE}
library(ggplot2)
library(GGally)
library(dplyr)
library(memisc)
library(DAAG)
library(Simpsons)
library(pander)
library(lazyeval)
library(gridExtra)
library(colorspace)
```
```{r}
#getwd()
wine <- read.csv('wineQualityReds.csv')

```

# Transform quality from an integer to a ordered factor
```{r}
wine$quality <- factor(wine$quality, ordered = T)
```
# create a rating varible
```{r}
wine$rating <- ifelse(wine$quality < 5, 'bad', ifelse(
  wine$quality < 7, 'average',  'good'))
wine$rating <- ordered(wine$rating,
                     levels = c('bad', 'average', 'good'))
```

#Univariate Plot Section
```{r}
print("Dataset variables")
pandoc.list(names(wine))

```

```{r}
print("Dataset structure")
str(wine)

```

```{r}
set.caption("Dataset summary")
pandoc.table(summary(wine))
```


```{r }
p1<-qplot(data = wine , x = quality)
p2<-qplot(data = wine , x = rating)
grid.arrange(p1,p2,ncol=1)

```


The high concentration of wines in both the cases is in the centre region as we can observe.

```{r }
ggplot(data=wine, aes(x=fixed.acidity)) +
  geom_histogram(binwidth = 1)+
  xlab('Fixed Value of Acidity')+
  ylab('Number of Samples')

  
```


There is a high concentration of wines with fixed.acidity close to 8 (the median) but there are also some outliers that shift the mean up to 9.2.

```{r, include = FALSE}
qplot(volatile.acidity, data = wine, binwidth = .01)
   
```
The distribution appears bimodal at 0.4 and 0.6 with some outliers in the higher ranges.

```{r}
qplot(citric.acid, data = wine, binwidth = .01)
```
This is strange distribution. Some percentage of wines do not have critic acid at all.
```{r}
qplot(residual.sugar, data = wine, binwidth = .1)
```
A high concentration of wines is around 2.2 (the median) with some outliers along the higher ranges.

```{r}
qplot(chlorides, data = wine, binwidth = .01)
```

A high concentration of wines is around 2.2 (the median) with some outliers along the higher ranges.

```{r}
qplot(free.sulfur.dioxide, data = wine, binwidth = 1)
```

The distributions peaks at around 7 and from then on resembles a long tailed distribution with very few wines over 60.

```{r}
qplot(total.sulfur.dioxide, data = wine, binwidth = 2)
```
The distributions peaks at around 7 and from then on resembles a long tailed distribution with very few wines over 150

```{r}
qplot(density, data = wine, binwidth = 0.001)
```

This distribution is almost symmetrical and pretty normal.

```{r}
qplot(pH, data = wine, binwidth = 0.1) 
  
```
pH looks normally distributed

```{r }
qplot(sulphates, data = wine, binwidth = .1)+
  scale_x_continuous(breaks=c(0.3,0.5,0.7,0.9,1.1,1.3,1.5,1.7,1.9,2.1))
  
```

A high concentration of wines is around 0.6 (the median) with some outliers along the higher ranges.There is a rapid increase and then long tailed distribution 

```{r }
qplot(alcohol, data = wine, binwidth = .1)
  
```
There is a rapid increase and then long tailed distribution as seen for sulphur dioxide.

#Univariate Analysis
There are 1599 wine samples in the datasetwith 14 variables(X, fixed.acidity ,volatile.acidity ,citric.acid ,residual.sugar ,chlorides ,free.sulfur.dioxide ,total.sulfur.dioxide ,density ,pH ,sulphates ,alcohol 
,quality ,rating.
The variables quality and rating, are ordered factor variables with the following levels.
worst->best
quality -> 5,6,7
rating -> bad,average,good

###Other observations:
The median quality is 6, which in the given scale (1-10) is a mediocre wine. The better wine in the sample has a score of 8, and the worst has a score of 3. The dataset is not balanced, that is, there are a more average wines than poor or excelent ones.

###What is/are the main feature(s) of interest in your dataset?
The main feature in the data is quality and i would want to knnow which features determine the quality of wines.

###What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The variable residual.sugar dictates how sweet a wine is and might also have an influence in taste.The variables related to acidity (fixed, volatile, citric.acid and pH) might explain some of the variance.

###Did you create any new variables from existing variables in the dataset?
I created a rating variable to improvise the future visualizations.

###Of the features you investigated, were there any unusual distributions? Did
you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
Citric.acid stood out from the other distributions. 8% of the wines did not have the citric acid concentration at all which is unexpected.

#Bivariate Plots Section
Correlation tabble:in order to calculate the correlation matrix we need to transform the data a little by removing the unwanted columns.Here we translated quality into a number.
```{r , include = FALSE}
#chain functions together %.%
c <- cor(
  wine %>%
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
     quality = as.numeric(quality)
    )
  )

emphasize.strong.cells(which(abs(c) > .3 & c != 1, arr.ind = TRUE))
pandoc.table(c)
```
Alcohol has negative correlation with density. This is expected as alcohol is less dense than water.
Volatile.acidity has a positive correlation with pH. This is unexpected as pH is a direct measure of acidity. Maybe the effect of a lurking variable?
Residual.sugar does not show correlation with quality. Free.sulfur.dioxide and total.sulfur.dioxide are highly correlated as expected.
Density has a very strong correlation with fixed.acidity. The variables that have the strongest correlations to quality are volatile.acidity and alcohol.
Let's use boxplots to further examine the relationship between some varibles and quality.

Summarising using a box plot
```{r , include = FALSE}
summarise_boxplot <- function(data, x, y) {
  pandoc.table(data %>%
  group_by_(x) %>%
  summarise_(
    mean = interp(~mean(v), v=as.name(y)),
    median = interp(~median(v), v=as.name(y))
    ), caption = paste("Summaries for", y, "grouped by", x))
}
```

###Relationship between qualitY and acid concentration

```{r}
ggplot(data=wine, aes(x=quality, y=fixed.acidity)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "fixed.acidity")

```


As the correlation table showed, fixed.acidity seems to have little to no effect on quality.

```{r}
ggplot(data=wine, aes(x=quality,y=volatile.acidity)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "volatile.acidity")
```

Quality seems to go up when volatile.acidity goes down. The higher ranges seem to produce more average and poor wines.

```{r, include = FALSE}
ggplot(data=wine, aes(x=quality, y=citric.acid)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "citric.acid")

```
Better wines tend to have higher concentration of citric acid.

```{r}
ggplot(data=wine, aes(x=quality, y=residual.sugar)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "residual.sugar")
```
Sugar seems to have little effect on perceived quality.

```{r}
ggplot(data=wine, aes( x=quality ,y=chlorides)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "chlorides")
  
```
Altough weakly correlated, a lower concentration of chlorides seem to produce better wines.

```{r}
ggplot(data=wine, aes(y=free.sulfur.dioxide, x=quality)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "free.sulfur.dioxide")

```

The ranges are really close to each other ,for little sulfur dioxide and we get a poor wine and for too much and we get an average wine.
```{r}
ggplot(data=wine, aes(y=total.sulfur.dioxide, x=quality)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "total.sulfur.dioxide")

```
The distribution is similar to free.sulphur.dioxide

```{r}
ggplot(data=wine, aes(y=density, x=quality)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "density")

```
Better wines tend to have lower densities, but this is probably due to the alcohol concentration. I wonder if density still has an effect if we hold alcohol constant.

```{r}
ggplot(data=wine, aes(y=pH, x=quality)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "pH")
```
Altough there is definitely a trend (better wines being more acid) there are some outliers.I wonder how the distribution of the different acids affects this.


###Relationship between pH and acid concentration


```{r}
ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.5) +
  scale_x_log10(breaks=seq(5,15,1)) +
  xlab("log10(fixed.acidity)") +
  geom_smooth(method="lm")
 

```

Negative correlation.
The fixed.acidity decreases with increase in pH.

```{r}
ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point() +
  scale_x_log10(breaks=seq(.1,1,.1)) +
  xlab("log10(volatile.acidity)") +
  geom_smooth(method="lm")

```
Positive correlation
The concentration of volatile.acidity is concentrated more between 0.4 and 0.8.The higher ranges tend to have less concentration

```{r}
ggplot(data = subset(wine, citric.acid > 0), aes(x = citric.acid, y = pH)) +
  geom_point() +
  scale_x_log10() +
  xlab("log10(citric.acid)") +
  geom_smooth(method="lm")

```
Negative correlation
The concentration of volatile.acidity is concentrated more between 0.10 and 0.19.The lower ranges tend to have less concentration

It is really strange that an acid concentration would have a positive correlation with pH.


We can investigate how much of the variance in pH these three acidity variables can explain using a linear model.

```{r, include = FALSE}
acidsLM <- lm(pH ~ 
                I(log10(citric.acid)) +
                I(log10(volatile.acidity)) +
                I(log10(fixed.acidity)),
                    data = subset(wine,citric.acid > 0))
summary(acidsLM)
df = data.frame(wine$quality )
df$predictions <- predict(acidsLM, wine)
df$error <- (df$predictions - wine$pH)/wine$pH



```

```{r}
ggplot(data=df, aes(x=wine.quality, y=error)) +
  geom_boxplot()
```

It seems the three acidity variables can only explain half the variance in PH. The mean error is specially bad on poor and on excellent wines. This leads me to believe that there are other component that affect acidity.

```{r}
ggplot(data=wine, aes(y=sulphates, x=quality)) +
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "sulphates")
```

Altough there are many outliers in the medium wines, better wines seem to have a higher concentration of sulphates.

```{r}
ggplot(data=wine, aes(y=alcohol, x=quality)) + 
  geom_boxplot()+
  summarise_boxplot(wine, "quality", "alcohol")
```
The correlation is clear here. With an increase in alcohol graduation we see an increase in the concentration of better graded wines. Given the high number of outliers it seems we cannot rely on alcohol alone to produce better wines. 

#Bivariate Ananlysis

###Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Fixed.acidity seems to have little to no effect on quality
Quality seems to go up when volatile.acidity goes down. The higher ranges seem to produce more average and poor wines.
Better wines tend to have higher concentration of citric acid.
Contrary to what I initially expected residual.sugar apparently seems to have little to no effect on perceived quality.
Altough weakly correlated, a lower concentration of chlorides seem to produce better wines.
Better wines tend to have lower densities.
In terms of pH it seems better wines are more acid but there were many outliers. Better wines also seem to have a higher concentration of sulphates.

###Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
I verified the strong relation between free and total sulfur.dioxide.
I also checked the relation between the acid concentration and pH. Of those, only volatile.acidity surprised me with a positive coefficient for the linear model.

###What was the strongest relationship you found?
The relationship between the variables total.sulfur.dioxide and free.sulfur.dioxide.

#Multivariate Plots

```{r}
ggplot(data = wine,
       aes(y = density, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()+
  theme_dark()
```

When we hold alcohol constant, there is no evidence that density affects quality which confirms our earlier suspicion.

```{r}
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  scale_color_brewer()+
  theme_dark()
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  facet_wrap(~rating) +
  scale_color_brewer()+
  theme_classic()+
  theme_dark()
```

For wines with high alcohol content, having a higher concentration of sulphates produces better wines.


```{r}
ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()+
  theme_dark()
ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating) +
  scale_color_brewer()+
  theme_classic()+
  theme_dark()
```

The reverse seems to be true for volatile acidity. Having less acetic acid on higher concentration of alcohol seems to produce better wines.

```{r}
ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()+
  theme_dark()
ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating) +
  scale_color_brewer()+
  theme_classic()+
  theme_dark()
```

Low pH and high alcohol concentration seem to be a good combination.

###Acid exploration
Using multivariate plots we should be able to investigate further the relationship between the acids and quality.

```{r}
ggplot(data = wine,
       aes(y = fixed.acidity, x = volatile.acidity,
           color = quality)) +
  geom_point() +
  scale_color_brewer()+
  theme_dark()

```

Almost no variance in the y axis compared to the x axis.


```{r}
ggplot(data = wine,
       aes(y = citric.acid, x = volatile.acidity,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating) +
  scale_color_brewer()+
  theme_dark()

```

High citric acid and low acetic acid seems like a good combination.

```{r}
ggplot(data = wine,
       aes(y = citric.acid, x = fixed.acidity,
           color = quality)) +
  geom_point() +
  scale_color_brewer()+
  theme_dark()
  cor.test(wine$citric.acid, wine$fixed.acidity) 
```

Altough there seems to a correlation between tartaric acid and citric acid concentrations, nothing stands out in terms of quality.

#Multivariate Analysis

###Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
High alcohol contents and high sulphate concentrations combined seem to produce better wines.

#Final Plots and Summary

###Plot 1
```{r}
ggplot(data=wine, aes(y=alcohol, x=quality)) + 
  geom_boxplot() +
  ylab("alcohol concentration (% by volume)") +
  xlab("quality factor of wine")
  ggtitle("Influence of alcohol on wine quality")
```

Description:
Alcohol has a big influence on wine. So its an important factor to look for the next time you go alcohol shopping as higher the amount of alcohol stronger is the wine.

###Plot 2
```{r}
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("pottasium sulphate (g/dm3)") +
  xlab("alcohol (% by volume)") +
  scale_color_brewer() +
  ggtitle("Alcohol and sulphates over wine quality")

```

Description:
High alcohol contents and high sulphate concentrations combined seem to produce better wines.



#Reflection

The wine data set contains information on the chemical properties of a selection of wines collected. It also includes wine ranking.
The first thing I noticed was the high concentration of wines in the middle ranges of the ranking, that is, average tasting wines.Out of the chemical varibles, the only one that stood out was the concentration of citric acid (variable name citric.acid). First thing i noticed was the high number of wines that had no citric.acid at all.Nonetheless this variable also showed a strange distribution with some peaks but showing an almost rectangular distribution specially in the 0-0.5 range.
All of the other variables showed either an normal or long tailed looking distribution.
After exploring the individual variables, I proceded to investigate the relationships between each input variable and the outcome variable quality.
The most promissing varibles were alcohol concentration, sulphates and the individual acid concentrations.
I also tried investigating the effect of each acid in the overall pH for the wine. I used scatterplots to explore the relationships graphically and also generated a linear model to check how much of pH the three variables accounted for.
The first surprise here was finding that the correlation between acetic acid concentration and pH was positive. I immediately suspected this was the result of some lurking variable Simpsons paradox.
The second finding was discovering that the concentration of the three acids only account for less than half of the variance in pH. I interpreted this as a sign that there more components affecting acidity that were not measured.
On the final part of the analysis I tried using multivariate plots to investigate if there were interesting combinations of variables that might affect quality. I also used a multivariate plot to confirm that density did not have an effect on quality when holding alcohol concentration constant.
I think it would be interesting to include each wine critic judgement as separate entry in the dataset. After all, each individual has a different taste and is subject to prejudice and other distorting factors. I believe that having this extra information would add more value to the analysis.

